Fixing Vulnerabilities Potentially Hinders Maintainability

### 2021-02-08 response/response.tex

\Us As we described in comment $2.19$, we manually inspected $25$ cases.
Many of those cases were changes of small size ($< 10$ LOCs) where
refactorings were usually improving names of variables, functions, cases of type
conversion, etc. Changes that have no impact on the metrics we are 
evaluating.

Our concern by presenting a total random baseline 
was that comparing regular changes with security changes with very different sizes---a scenario 
that could happen---was unfair. Thus, we chose to create this baseline.
However, we also curated a total random baseline in the past and we now 
present both baselines in the paper.

\begin{revcomment}[2.21]
    Section 4\\
    \begin{itemize}
        \item RQ3 is answered to succinctly and with too little 
        detail to be convincing to me that the differences between 
        security and regular changes is different.  
    \end{itemize}
\end{revcomment}

\Us We improved the discussion of the comparison between 
the results for security patches and regular changes. We now
consider two baselines: \emph{size-baseline}, where 
we collected random changes with the same size as security 
patches; and, \emph{random-baseline}, where we collected
random changes without any restrictions. The overall results
show that regular changes are less prone to hinder software 
maintainability than security patches. Security
patches results have more cases with negative 
impact on software maintainability than any baseline 
of regular changes. By providing results 
for both baselines, we now clearly see that the limitation 
found for the \emph{size-baseline} does not change 
the need to pay more attention to security patches 
in software maintenance tasks.

@@

\begin{revcomment}[2.22]
    Section 5\\
    \begin{itemize}
        \item Line 32 - since RQ3 was not convincing, 
        I don't feel a separate effort for maintainable security 
        is justified as separate from general focus on maintainability. 
    \end{itemize}
\end{revcomment}

\Us As explained in the previous response, security patches 
hinder more software maintainability than regular changes with size 
restriction or no size restriction. Thus, we argue that
special attention should be given to maintainable security
when teaching software maintainability.

### 2021-02-08 paper/emse20.tex
adding RQ3

This research performs the following main contributions:
%
\begin{itemize}
	\vspace{-0.3em}
  \item Evidence that supports the trade-off between security and 
  maintainability: developers may be hindering software 
  maintainability while patching vulnerabilities.
	\item An empirical study on the impact of security patches on 
	software maintainability (per guideline, severity, weakness and 
	programming language).
	\item A replication package with the scripts and data created to 
	perform the empirical evaluation, for reproducibility. Available 
	online: \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: Section~\ref{sec:motivation} 
introduces an example of a security patch of a known vulnerability 
found in the protocol implementation of 
OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at 
\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
Section~\ref{sec:methodology} describes the methodology used to 
answer the research questions; Section~\ref{sec:results} presents 
the results and discusses their implications; 
Section~\ref{sec:implications} elaborates on the implications
developers should consider in the future; Section~\ref{sec:threats} 
enumerates the threats to the validity of this study; 
Section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, 
Section~\ref{sec:conclusions} concludes the main findings and 
elaborates on future work.

@@

\textit{\textbf{RQ3: What is the impact of security changes versus regular 
changes on the maintainability of open-source software?}}
The impact of security and regular changes on software maintainability 
is presented in Figure~\ref{fig:secvsreg}. In this section, 
we present a comparison of security patches with 
two different baselines of regular changes: \emph{random-baseline},
a dataset of random regular changes;
and \emph{size-baseline}, a dataset of  
regular changes with the same size as security 
patches. Our hypothesis in this research question is
that \emph{Security patches hinder more software maintainability and 
need more attention than regular changes}.

We have seen, previously, a deterioration on software maintainability 
when patching vulnerabilities: $41.9\%$ ($406$) suffered a 
negative impact, $38.7\%$ ($375$) remained the same and $19.4\%$ 
($188$) increased software maintainability. For regular changes, 
we observe that the maintainability decreases in $27.0\%$ ($262$) 
and increases in $30.5\%$ ($295$) of the cases. But in contrast to 
security patches, the maintainability of regular changes remains the 
same in $42.5\%$ ($412$) of the cases and perform regular
changes has a more positive impact than negative on  
maintainability.

@@

Results show that regular changes are less prone to hinder the software maintainability of open-source software.   
We manually inspected a considerable amount of cases from
the distribution of the regular changes where maintainability 
increases and found that identifying regular commits as similar in 
size as the security-related commit is limiting the type of regular 
commits being randomly chosen (e.g, input 
patches, variables or functions, type conversion, etc).  
We presume that this phenomenon lead
to the significant number of cases where there is no impact
on the software maintainability.

\textbf{Summary:} Security-related commits are observed to harm software 
maintainability. In contrast, one cannot say the same about
regular changes, in which \textbf{no significant differences are observed}. 
Thus, we urge the importance of adopting maintainability practices while 
applying security patches.

>>>>

This research performs the following main contributions:
%
\begin{itemize}
  \item Evidence that supports the trade-off between security and 
  maintainability: developers may be hindering software 
  maintainability while patching vulnerabilities.
	\item An empirical study on the impact of security patches on 
	software maintainability (per guideline, severity, weakness and 
	programming language).
	\item A replication package with the scripts and data created to 
	perform the empirical evaluation, for reproducibility. Available 
	online: \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: Section~\ref{sec:motivation} 
introduces an example of a security patch of a known vulnerability 
found in the protocol implementation of 
OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at 
\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
Section~\ref{sec:methodology} describes the methodology used to 
answer the research questions; Section~\ref{sec:results} presents 
the results and discusses their implications; Section~\ref{sec:lessons}
present the lessons learned of this study;
Section~\ref{sec:implications} elaborates on the implications
developers should consider in the future; Section~\ref{sec:threats} 
enumerates the threats to the validity of this study; 
Section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, 
Section~\ref{sec:conclusions} concludes the main findings and 
elaborates on future work.

@@

\textit{\textbf{RQ3: What is the impact of security changes versus regular 
changes on the maintainability of open-source software?}}
The impact of security and regular changes on software maintainability 
is presented in Figure~\ref{fig:secvsreg}. In this section, 
we present a comparison of security patches with 
two different baselines of regular changes: 
\emph{size-baseline}, a dataset of  
random regular changes with the same size as security 
patches becausecomparing changes with considerable different sizes may be unfair; and, \emph{random-baseline},
a dataset of random regular changes. Our hypothesis in this research question is
that \emph{Security patches hinder more software maintainability and 
need more attention than regular changes}.

We have seen, previously, a deterioration on software maintainability 
when patching vulnerabilities: $41.9\%$ ($406$) of patches suffered a 
negative impact, $38.7\%$ ($375$) of patches remained the same and $19.4\%$ 
($188$) of patches increased software maintainability. For regular changes,
when considering the size of the changes (\emph{size-basline}),
we observe that the maintainability decreases in $27.0\%$ ($262$) 
and increases in $30.5\%$ ($295$) of the cases. But in contrast to 
security patches, the maintainability of regular changes remains the 
same in $42.5\%$ ($412$) of the cases and perform regular
changes has a more positive impact than negative on  
maintainability. Fully regular changes (\emph{random-baseline}) 
are less prone to hinder software maintainability than 
security changes. About $34.4\%$ ($333$) of the regular changes
have a negative impact on software maintainability compared with 
the $41.9\%$ of security patches.

@@

Overall, the results for both baselines, show that regular 
changes are less prone to hinder the software 
maintainability of open-source software.   
We manually inspected a considerable amount of cases from
the distribution of regular changes that considers size (\emph{size-baseline}), 
where the impact on maintainability 
increases, and, found that identifying regular changes with the same 
size as the security-related commit is limiting the type of regular 
commits being randomly chosen: input 
patches, variables or functions, type conversion, i.e., changes
with no impact on the software metrics BCh evaluates.
We presume that this phenomenon lead
to the significant number of cases where there is no impact
on the software maintainability. On the other hand, 
identifying regular changes without any restrictions (\emph{random-baseline})
shows that regular changes have less negative impact 
on software maintainability compared with security patches
with statistical significance ($p = 5.34$x$10^{-8}$) and that 
special attention should be given to security patches.

\textbf{Summary:} Security-related commits are observed to harm software 
maintainability. While regular changes with no characteristics restrictions 
show that regular changes are less prone to harm software maintainability.
Thus, we urge the importance of adopting maintainability practices while 
applying security patches.
